Efficient and Scalable Bipartite Matching with Fast Beta Linkage (fabl)

### 2023-07-30 ba_resubmit_template_files/main.tex
adding in supp to main file and checking for typos.
fix typos using antidote 11.

Having defined comparison vectors, we now turn to the \cite{fellegi_theory_1969} model for record linkage.  We begin with notation for defining linkages. Under bipartite matching, the set of matches across $A$ and $B$ can be represented in two equivalent ways. First, we may use a matrix $\Delta \in \{0, 1\}^{n_An_B}$, where
\begin{align}
\Delta_{ij} =
\begin{cases}
1, & \text{if records}\;  A_i \; \text{and}\; B_j \; \text{refer to the same entity}; \\
0, & \text{otherwise}.\\
\end{cases}
\end{align}
This sparse matrix representation can become cumbersome for large linkage tasks. More compactly, bipartite matching also can be viewed as a labeling $\bm{Z} = (Z_1, \ldots, Z_{n_B})$ for the records in $B$ such that
\begin{align}
Z_{j} =
\begin{cases}
i, & \text{if records}\;  A_i \; \text{and}\; B_j  \; \text{refer to the same entity}; \\
n_A + j, & \text{if record}\;  B_j \; \text{does not have a match in}\; A. \\
\end{cases}
\end{align}
We can go back and forth between the two using $\Delta_{ij} = I(Z_j = i),$ where $I(\cdot) = 1$ when the expression inside the parentheses is true, and $I(\cdot) = 0$ otherwise. When presenting the models, we use both representations for convenience.

@@

Finally, to obtain an estimate $\hat{\bm{Z}}$ of the linkage structure, we use the loss functions and Bayes estimate from \cite{sadinle_bayesian_2017}. Since (\ref{eqn:fast_beta_prior}) does not strictly enforce one-to-one matching, it is possible for this Bayes estimate to link multiple records in $B$ to one record in $A$. To obtain a Bayes estimate that fulfills the bipartite requirement, we minimize the expected loss subject to the constraint that $\hat{Z}_j \neq \hat{Z}_{j'}$ for all $j \neq j'$. See Supplement \ref{bayes-estimate} for details redarding the initial Bayes estimate and this post-processing procedure.

@@

\begin{lemma}
	Let $n_A$ and $n_B$ be the number of records in files $A$ and $B$, respectively. Let $F$ be the number of fields used for comparisons across records, and $P$ be the number of patterns that comparison vectors exhibit in $A \times B$. We assume $C$ cores available for parallelization and a Gibbs sampler with $T$ iterations. Then, the overall computational complexity of \texttt{fabl} with hashing is $O(F n_A n_B /C) + O(T n_B P / C)$.
	\label{lemma:fabl}
\end{lemma}
\begin{proof}
	We consider two steps: constructing the comparison vectors and the Gibbs sampler. The computational complexity of all pairwise comparisons across $A$ and $B$ is $O(F n_A n_B)$. The hashing procedure for all pairwise comparisons is also $O(F n_A n_B)$. With $C$ processors available, we can split these computations across $C$ equally sized partitions and compute these comparisons in parallel, so the complexity becomes $O(F n_A n_B /C)$. There are then trivial computational costs associated with synthesizing summary statistics across these partitions. We note that this contribution to computational complexity applies for all versions of \cite{fellegi_theory_1969} algorithms, unless they use blocking to reduce the comparison space.

	Without hashing, the computational complexity of updating the $\bm{m}$ and $\bm{u}$ parameters is $O(F n_A n_B)$. However, by doing calculations over the agreement patterns rather than the individual records, hashing reduces the overall complexity to $O(P)$. The complexity of updating $\bm{Z}$ sequentially at the record level as in \cite{sadinle_bayesian_2017} is $O(n_A n_B)$. With hashing, we first sample the agreement pattern of the match with complexity $O(n_B P)$, and then we sample the record exhibiting that pattern with complexity $O(n_B)$. Thus, the complexity of sampling $\bm{Z}$ in a single iteration is $O(n_B P)$. Since $P << n_A$ in most applications, we have reduced the complexity of sampling $\bm{Z}$ from $O(F n_A n_B)$ under \texttt{BRL} to $O(n_B P)$ under \texttt{fabl}. With parallelization, this complexity is further reduced to $O(n_B P /C)$, and so the entire Gibbs sampler has complexity $O(T n_B P / C)$
	In summary, the total computational complexity for \texttt{fabl} is $O(F n_A n_B / C) + O(T n_B P / C).$
\end{proof}

@@

		We then conduct hashing, obtain the compressed $\tilde{\gamma}^{ab}$, and delete the memory intensive matrix $\gamma^{ab}$ before continuing with the next batch of data. In detail, we calculate
\begin{subequations}
	\begin{align}
		r_{p_j}^{ab} &= \{i \in 1, \dots, n_A| \gamma_{ij} = h_p, B_j \in B^b\}, \label{r_{p_j}^{ab}} \\
		N_{p_j}^{ab} &= |r_{p_j}^{ab}|. \label{N_{p_j}^{ab}}
	\end{align}
\end{subequations}
These can computed sequentially or in parallel. Summary statistics from each pairwise batch comparison can be synthesized to recover summary statistics for the full comparison matrix $\gamma$ through
\begin{subequations}
	\begin{align}
		r_{p_j} &= (r_{p_j}^{11}, \ldots, r_{p_j}^{ t_A t_B}) \text{ for } a = 1, \ldots, t_A \text{ and } b = 1, \ldots, t_B, \label{r_{p_j}} \\
		N_{p_j} &= \sum_{a = 1}^{t_A} \sum_{b = 1}^{t_B} N_{p_j}^{ab}.\label{N_{p_j}}	\end{align}
\end{subequations}

@@

	In Figure \ref{fig:speed2}, where we fix $n_B = 500$, we see near linear growth for the run-time under \texttt{BRL} as $n_A$ increases, and much more static run-time under \texttt{fabl}. The slight increases in run-time for \texttt{fabl} are due primarily to the hashing step, which again can be run in parallel for large data. To illustrate that these trends are generalizeable to other specifications of the comparison vectors, we have included the run-time results for an additional simulation study, under different comparison vector settings, in Supplement \ref{app:appendix-speed}.

@@

	The country of El Salvador was immersed in civil war from 1980 to 1991. We are interested in estimating the total number of individuals killed in the war. We utilize lists of documented deaths from the war, one collected by El Rescate - Tutela Regal (ERTL) and another from the Salvadoran Human Rights Commission (CDHES, by its acronym in Spanish).\footnote{We thank the Human Rights Data Analysis Group (HRDAG) for granting access to these data.} The ERTL dataset comprises digitized denunciations published throughout the conflict, and the CDHES dataset comprises killings reported directly to the organization \citep{howland2008rescate, ball2000salvadoran, green2019civilian}. The ERTL required additional investigation before recording denunciations as human rights abuses, and reports to the CDHES were made shortly after the events occurred; thus, both data files are thought to be fairly reliable. When estimating the total number of individuals killed, one cannot simply sum the numbers recorded by each organization, as it is likely that the same individuals are recorded in multiple casualty lists. Instead, record linkage techniques must be used to merge data files before analyzing the data \citep{lum2013applications}.

@@

	Therefore, we perform our analysis with the agreement levels for each field according to Table \ref{Tab:el_salvador_cutoffs_2}. Among the 432 possible agreement patterns, 159 are realized in the data. With this revised comparison specification, \texttt{fabl} runs in 109 seconds, approximately 3 times faster than the \texttt{BRL} run time of 313 seconds.  The estimates of the $\bm{m}$ parameters under each method are similar, as shown in Figure \ref{fig:m-and-u}. Estimates of the $\bm{u}$ parameters are indistinguishable, and thus omitted. Traceplots for parameters of interest are provided in Supplement \ref{app:appendix-es}.

@@

	\begin{table}
		\centering
		\begin{tabular}[t]{lllll}
			\multicolumn{2}{c}{ } & \multicolumn{3}{c}{Level of Disagreement} \\
			\cline{3-5}
			Fields & Similarity & 1 & 2 & 3\\
			\hline
			Sex & Binary & Agree & Disagree & \\
			Year of Birth & Binary & Agree & Disagree & \\
			Month of Birth & Binary & Agree & Disagree & \\
			Day of Birth & Binary & Agree & Disagree & \\
			Location & Custom & Same State and Office & Same State & Otherwise \\
			\hline
		\end{tabular}
		\caption{Construction of comparison vectors for NTLCS data.}\label{Tab:nltcs-comparisons}
	\end{table}

	We link records using sex, location, and day, month, and year of birth using the criteria shown in Table \ref{Tab:nltcs-comparisons}. Storing $\gamma$ constructed through three comparison scores for each of $20485 (17466) \approx$ 400 million record pairs would require approximately 8GB of memory. Standard settings on a 16GB personal computer do not allow storage of an object of this size, and thus \texttt{BRL} is unable to perform this linkage task on such a machine. However, through the method described in Section \ref{scaling},  we perform 30 smaller comparison tasks, using $t_A = 1$ and $t_B = 30$. We conduct linkage with all record indices recorded and also with SEI using $S=10$, and obtain identical results. The hashed $\tilde{\gamma}$ without SEI is about 2.2 GB, and with SEI, it is about 760 MB. Constructing the comparisons sequentially took approximately 40 minutes, which could be reduced considerably through parallel computing.

>>>>

Having defined comparison vectors, we now turn to the \cite{fellegi_theory_1969} model for record linkage.  We begin with notation for defining linkages. Under bipartite matching, the set of matches across $A$ and $B$ can be represented in two equivalent ways. First, we may use a matrix $\Delta \in \{0, 1\}^{n_An_B}$, where
\begin{align}
\Delta_{ij} =
\begin{cases}
1, & \text{if records}\;  A_i \; \text{and}\; B_j \; \text{refer to the same entity}; \\
0, & \text{otherwise}.\\
\end{cases}
\end{align}
This sparse matrix representation can become cumbersome for large linkage tasks. More compactly, bipartite matching can also be viewed as a labeling $\bm{Z} = (Z_1, \ldots, Z_{n_B})$ for the records in $B$ such that
\begin{align}
Z_{j} =
\begin{cases}
i, & \text{if records}\;  A_i \; \text{and}\; B_j  \; \text{refer to the same entity}; \\
n_A + j, & \text{if record}\;  B_j \; \text{does not have a match in}\; A. \\
\end{cases}
\end{align}
We can go back and forth between the two using $\Delta_{ij} = I(Z_j = i),$ where $I(\cdot) = 1$ when the expression inside the parentheses is true, and $I(\cdot) = 0$ otherwise. When presenting the models, we use both representations for convenience.

@@

Finally, to obtain an estimate $\hat{\bm{Z}}$ of the linkage structure, we use the loss functions and Bayes estimate from \cite{sadinle_bayesian_2017}. Since (\ref{eqn:fast_beta_prior}) does not strictly enforce one-to-one matching, it is possible for this Bayes estimate to link multiple records in $B$ to one record in $A$. To obtain a Bayes estimate that fulfills the bipartite requirement, we minimize the expected loss subject to the constraint that $\hat{Z}_j \neq \hat{Z}_{j'}$ for all $j \neq j'$. See Supplement \ref{bayes-estimate} for details regarding the initial Bayes estimate and this post-processing procedure.

@@

\begin{lemma}
	Let $n_A$ and $n_B$ be the number of records in files $A$ and $B$, respectively. Let $F$ be the number of fields used for comparisons across records, and $P$ be the number of patterns that comparison vectors exhibit in $A \times B$. We assume $C$ cores available for parallelization and a Gibbs sampler with $T$ iterations. Then, the overall computational complexity of \texttt{fabl} with hashing is $O(F n_A n_B /C) + O(T n_B P / C)$.
	\label{lemma:fabl}
\end{lemma}
\begin{proof}
	We consider two steps: constructing the comparison vectors and the Gibbs sampler. The computational complexity of all pairwise comparisons across $A$ and $B$ is $O(F n_A n_B)$. The hashing procedure for all pairwise comparisons is also $O(F n_A n_B)$. With $C$ processors available, we can split these computations across $C$ equally sized partitions and compute these comparisons in parallel, so the complexity becomes $O(F n_A n_B /C)$. There are then trivial computational costs associated with synthesizing summary statistics across these partitions. We note that this contribution to computational complexity applies for all versions of \cite{fellegi_theory_1969} algorithms unless they use blocking to reduce the comparison space.

	Without hashing, the computational complexity of updating the $\bm{m}$ and $\bm{u}$ parameters is $O(F n_A n_B)$. However, by doing calculations over the agreement patterns rather than the individual records, hashing reduces the overall complexity to $O(P)$. The complexity of updating $\bm{Z}$ sequentially at the record level as in \cite{sadinle_bayesian_2017} is $O(n_A n_B)$. With hashing, we first sample the agreement pattern of the match with complexity $O(n_B P)$, and then we sample the record exhibiting that pattern with complexity $O(n_B)$. Thus, the complexity of sampling $\bm{Z}$ in a single iteration is $O(n_B P)$. Since $P << n_A$ in most applications, we have reduced the complexity of sampling $\bm{Z}$ from $O(F n_A n_B)$ under \texttt{BRL} to $O(n_B P)$ under \texttt{fabl}. With parallelization, this complexity is further reduced to $O(n_B P /C)$, and so the entire Gibbs sampler has complexity $O(T n_B P / C).$
	In summary, the total computational complexity for \texttt{fabl} is $O(F n_A n_B / C) + O(T n_B P / C).$
\end{proof}

@@

		We then conduct hashing, obtain the compressed $\tilde{\gamma}^{ab}$, and delete the memory intensive matrix $\gamma^{ab}$ before continuing with the next batch of data. In detail, we calculate
\begin{subequations}
	\begin{align}
		r_{p_j}^{ab} &= \{i \in 1, \dots, n_A| \gamma_{ij} = h_p, B_j \in B^b\}, \label{r_{p_j}^{ab}} \\
		N_{p_j}^{ab} &= |r_{p_j}^{ab}|. \label{N_{p_j}^{ab}}
	\end{align}
\end{subequations}
These can be computed sequentially or in parallel. Summary statistics from each pairwise batch comparison can be synthesized to recover summary statistics for the full comparison matrix $\gamma$ through
\begin{subequations}
	\begin{align}
		r_{p_j} &= (r_{p_j}^{11}, \ldots, r_{p_j}^{ t_A t_B}) \text{ for } a = 1, \ldots, t_A \text{ and } b = 1, \ldots, t_B, \label{r_{p_j}} \\
		N_{p_j} &= \sum_{a = 1}^{t_A} \sum_{b = 1}^{t_B} N_{p_j}^{ab}.\label{N_{p_j}}	\end{align}
\end{subequations}

@@

	In Figure \ref{fig:speed2}, where we fix $n_B = 500$, we see near linear growth for the run-time under \texttt{BRL} as $n_A$ increases, and much more static run-time under \texttt{fabl}. The slight increases in the run-time for \texttt{fabl} are due primarily to the hashing step, which again can be run in parallel for large data. To illustrate that these trends are generalizable to other specifications of the comparison vectors, we have included the run-time results for an additional simulation study, under different comparison vector settings, in Supplement \ref{app:appendix-speed}.

@@

	The country of El Salvador was immersed in civil war from 1980 to 1991. We are interested in estimating the total number of individuals killed in the war. We utilize lists of documented deaths from the war, one collected by El Rescate - Tutela Regal (ERTL) and another from the Salvadoran Human Rights Commission (CDHES, by its acronym in Spanish).\footnote{We thank the Human Rights Data Analysis Group (HRDAG) for granting access to these data.} The ERTL data set comprises digitized denunciations published throughout the conflict, and the CDHES data set comprises killings reported directly to the organization \citep{howland2008rescate, ball2000salvadoran, green2019civilian}. The ERTL required additional investigation before recording denunciations as human rights abuses, and reports to the CDHES were made shortly after the events occurred; thus, both data files are thought to be fairly reliable. When estimating the total number of individuals killed, one cannot simply sum the numbers recorded by each organization, as it is likely that the same individuals are recorded in multiple casualty lists. Instead, record linkage techniques must be used to merge data files before analyzing the data \citep{lum2013applications}.

@@

	Therefore, we perform our analysis with the agreement levels for each field according to Table \ref{Tab:el_salvador_cutoffs_2}. Among the 432 possible agreement patterns, 159 are realized in the data. With this revised comparison specification, \texttt{fabl} runs in 109 seconds, approximately 3 times faster than the \texttt{BRL} run-time of 313 seconds.  The estimates of the $\bm{m}$ parameters under each method are similar, as shown in Figure \ref{fig:m-and-u}. Estimates of the $\bm{u}$ parameters are indistinguishable, and thus omitted. Traceplots for parameters of interest are provided in Supplement \ref{app:appendix-es}.

@@

	\begin{table}
		\centering
		\begin{tabular}[t]{lllll}
			\multicolumn{2}{c}{ } & \multicolumn{3}{c}{Level of Disagreement} \\
			\cline{3-5}
			Fields & Similarity & 1 & 2 & 3\\
			\hline
			Sex & Binary & Agree & Disagree & \\
			Year of Birth & Binary & Agree & Disagree & \\
			Month of Birth & Binary & Agree & Disagree & \\
			Day of Birth & Binary & Agree & Disagree & \\
			Location & Custom & Same State and Office & Same State & Otherwise \\
			\hline
		\end{tabular}
		\caption{Construction of comparison vectors for NLTCS data.}\label{Tab:nltcs-comparisons}
	\end{table}

	We link records using sex, location, and day, month, and year of birth using the criteria shown in Table \ref{Tab:nltcs-comparisons}. Storing $\gamma$ constructed through three comparison scores for each of $20485 (17466) \approx$ 400 million record pairs would require approximately 8 GB of memory. Standard settings on a 16 GB personal computer do not allow storage of an object of this size, and thus \texttt{BRL} is unable to perform this linkage task on such a machine. However, through the method described in Section \ref{scaling},  we perform 30 smaller comparison tasks, using $t_A = 1$ and $t_B = 30$. We conduct linkage with all record indices recorded and also with SEI using $S=10$, and obtain identical results. The hashed $\tilde{\gamma}$ without SEI is about 2.2 GB, and with SEI, it is about 760 MB. Constructing the comparisons sequentially took approximately 40 minutes, which could be reduced considerably through parallel computing.
